---
title: "Demographic and Economic Analysis of European Countries"
author: "Author: Klaudia Sołek"
output:
  html_document:
    css: styles.css
---

```{r}
library(gapminder)
library(dplyr)
library(ggplot2)
library(gcookbook)
library(MASS)
```

# Data "gapminder" used for the project, their structure, and basic statistics

```{r}
gapminder
str(gapminder)
summary(gapminder)
```

# Modification data

```{r}
gapminder_modified <- gapminder  
gapminder_modified$lifeExp <- round(gapminder$lifeExp, digits = 0)
gapminder_modified$gdpPercap <- round(gapminder$gdpPercap, digits = 1)
gapminder_modified
```
# Average life expectancy by continent in 2007 

```{r}
data_2007 <- subset(gapminder_modified, year == 2007)

avg_lifeExp <- data_2007 %>%
  group_by(continent) %>%
  summarise(mean_lifeExp = mean(lifeExp))

print(avg_lifeExp)
```

# A chart depicting how the population of Greece changed over the years from 1952 to 2007

```{r}
countries <- gapminder_modified %>%
    mutate(pop_milions = pop/1000000)

Greece <- countries %>%
  filter(country == "Greece") %>%
  arrange(year)

plot(Greece$year, Greece$pop_milions, type = "b",lty = 2,col = "darkviolet", lwd = 3,pch = 2, cex = 1.25,
     main = "Population in relation to the year",
     sub = "Data for Greece from 1952 to 2007",
     xlab = "year",
     ylab = "population [m]")
```

# A bar chart comparing the population change of Greece, Hungary and Bosnia and Herzegovina

```{r}
E2007 <- countries %>%
  filter(country %in% c("Greece", "Bosnia and Herzegovina", "Hungary"))
```

```{r}
ggplot(E2007, aes(x = year, y = pop_milions, fill = country)) + 
  geom_col(width = 2, position = position_dodge(3)) +
  labs(x = "year", y = "population [m]")
```

# Population in european countries in 2007

```{r}
europe_2007 <- countries %>%
  filter(year == 2007, continent == "Europe") %>%
  arrange(country)
```

```{r}
ggplot(europe_2007, aes(x = pop_milions, y = country)) +
  geom_segment(aes(yend = country), xend = 0, colour = "black") +
  geom_point(size = 3, colour = "blue") +
  labs(x = "population [m]")
```

# A scatter plot showing the change in GDP in Greece

```{r}
ggplot(Greece, aes(x = year, y = gdpPercap)) +
  geom_point(size=4,colour="green") +
  labs(x = "year", y = "GDP")
```

# Change in Greece's GDP compared to it's neighbors - Albania and Bulgaria

```{r}
Albania <- countries %>%
  filter(country == "Albania") %>%
  arrange(year)
Bulgaria <- countries %>%
  filter(country == "Bulgaria") %>%
  arrange(year)

par(mfcol = c(1,1))
plot(Greece$year,Greece$gdpPercap, type = "l", lty = 1, lwd = 3, col = "deeppink",   
     main = "Change in GDP", xlab = "year", ylab = "GDP value")
        lines(Albania$year,Albania$gdpPercap, col = "green4", type = "l", lty = 1, lwd = 2) 
        lines(Bulgaria$year,Bulgaria$gdpPercap, col = "darkblue", type = "l", lty = 1, lwd = 2)
        legend(x = 1951, y = 27000, legend=c("Greece", "Bulgaria", "Albania"),
               col=c("deeppink", "darkblue", "green4"), lty = 1)
```

# A bar chart comparing the GDP of Greece, Hungary and Bosnia and Herzegovina

```{r}
ggplot(E2007, aes(x = year, y = gdpPercap, fill = country)) +
  geom_col(colour = "black") +
  scale_fill_brewer(palette = "Pastel1")
```

# A chart showing GDP and population levels in european countries in 2007

```{r}
europe_2007$pop_millions <- europe_2007$pop/1000000

plot(europe_2007$pop_millions, europe_2007$gdpPercap, lwd = 1,
     main = "GDP vs. population of european countries",
     xlab = "Population [m]", ylab = "GDP")

text(europe_2007$pop_millions, europe_2007$gdpPercap, europe_2007$country, cex = 0.65, pos = 4, col = "blue")
```

# A bar chart depicting the change in life expectancy in Greece

```{r}
ggplot(Greece, aes(x = year, y = lifeExp)) +
  geom_col(fill = "lightblue", colour = "black") +
  geom_text(aes(label = lifeExp), vjust = 1.5, colour = "darkblue")
```

# A box plot representing life expectancy in Greece, Albania, and Bulgaria

```{r}
par(mfcol = c(1,3))
boxplot(Greece$lifeExp,col = "darkgreen", main="Greece", ylab="life expectancy")
boxplot(Albania$lifeExp,col = "darkblue", main="Albania", ylab="life expectancy")
boxplot(Bulgaria$lifeExp,col = "red", main="Bulgaria", ylab="life expectancy")
```

# A scatter plot of life expectancy of Europeans in 2007

```{r}
ggplot(europe_2007, aes(x = lifeExp, y = country)) +
  geom_point(size = 2, colour = "red") +
  labs(x = "life expectancy")
```





